### 2016-07-18 appendix.tex
Final submission manuscript

\section{HTTP Image Fetch Setup (Optional)}
\label{sec:http}

\input{testbed_http}

@@

Also, when creating the persistent home directory partition in
Section~\ref{sec:persistent_directories}, for recreating the loopback devices
the commands are:

>>>>

% \section{HTTP Image Fetch Setup (Optional)}
% \label{sec:http}

% \input{testbed_http}

@@

Also, when creating the persistent home directory partition in
Section~\ref{sec:non_persistent_directories}, for recreating the loopback
devices the commands are:

### 2016-07-18 cross_compilation.tex
Final submission manuscript

\subsection{Toolchain Setup}
To compile in a given architecture that is aimed for a different one,
the testbed administrator needs to install a toolchain in its \ac{PC}.
The toolchain is mandatory due to the different processor architectures
where the source can be compiled from.
Given that compiling a toolchaing can be an ardous task, we get the toolchain
recommended for the \ac{ARM} architecture of the \ac{Raspi}s.
This toolchain is available from \cite{2016rpitoolchain} and
it already contains the binaries for different compilers based on
\texttt{gcc 4.9}. In what follows, we extract the binaries adjusting
them to our coding style and compiling convention. For this, we use the
\texttt{\$TOOLCHAIN} directory as the working directory. However,
the testbed administrator may choose some other working directory of
its preference if desired. First, we create the toolchain directory:

\begin{lstlisting}[]
$ export TOOLCHAINDIR="${HOME}/toolchains"
$ mkdir -p $TOOLCHAINDIR
$ cd $TOOLCHAINDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

Instead of calling the ARM cross compiler using its full path, we make
the binaries accessible from the command shell systemwide. A way to do
this is by adding the following commands in the \texttt{\${HOME}/.profile}
as follows:

@@

To update the \texttt{PATH} variable and the \texttt{.profile}, we use the
\texttt{source} command for the changes take effect in the administrator
system:
\begin{lstlisting}[]
$ source ${HOME}/.profile
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\${CODEDIR}/hello\_world.cpp}@>
<@\textcolor{gray}{
-----------------------------------------------}
\Reactivatenumber @>
#include <iostream>

>>>>

\subsection{Toolchain Setup}
To compile in a given architecture that is aimed for a different one,
the testbed administrator needs to install a toolchain in its \ac{PC}.
The toolchain is mandatory due to the different processor architectures
where the source can be compiled from.
Given that compiling a toolchaing can be an ardous task, we get the toolchain
recommended for the \ac{ARM} architecture of the \ac{Raspi}s.
This toolchain is available from \cite{2016rpitoolchain} and
it already contains the binaries for different compilers based on
\texttt{gcc 4.9}. In what follows, we extract the binaries adjusting
them to our coding style and compiling convention. For this, we use the
\texttt{\$\{TOOLCHAIN\}} directory as the working directory. However,
the testbed administrator may choose some other working directory of
its preference if desired. First, we create the toolchain directory:

\begin{lstlisting}[]
$ export TOOLCHAINDIR="${HOME}/toolchains"
$ mkdir -p ${TOOLCHAINDIR}
$ cd ${TOOLCHAINDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

Instead of calling the ARM cross compiler using its full path, we make
the binaries accessible from the command shell systemwide. A way to do
this is by adding the following commands in the \texttt{\$\{HOME\}/.profile}
as follows:

@@

To update the \texttt{\$\{PATH\}} variable and the \texttt{.profile}, we
use the \texttt{source} command for the changes take effect in the
administrator system:
\begin{lstlisting}[]
$ source ${HOME}/.profile
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{CODEDIR\}/hello\_world.cpp}@>
<@\textcolor{gray}{
---------------------------------------------------}
\Reactivatenumber @>
#include <iostream>

### 2016-07-18 image_setup.tex
Final submission manuscript

To get started, we first need to install an \ac{OS} that works properly
on all the \ac{Raspi} models. We will download and setup the image in
the testbed administrator \ac{PC} using a Debian-based distribution. We
use the popular Debian-based Raspbian Linux~\cite{raspbian} given that is
the recommended and default \ac{OS} for the \ac{Raspi}. Raspbian is made
available in two bundles: Raspbian and Raspbian lite. The difference
between the two, is that Raspbian contains a pre-installed desktop environment
for user interaction and Raspbian lite by default only permits to interact
through a command shell. Given that the \ac{Raspi}s in our testbed are not
connected to monitors, we decide to work with Raspbian lite. Still, if later
desired, a desktop environment can be installed using the package manager
as it will be shown in the configuration section.

% The overall procedure to customize the official Raspbian lite image
% are:
% \begin{enumerate}
%     \item Download Raspbian lite
%     \item Alter Raspbian lite. e.g. browse, modify, add, and delete files
%         in the official Raspbian lite image
%     \item Change root, i.e. change root filesystem into the Raspbian lite
%         image to update and install software packages
%     \item Write image to memory cards
% \end{enumerate}

%\subsection{Download Raspbian lite}
%To download the latest Raspbian lite, we go to the url
%\url{http://downloads.raspberrypi.org/raspbian_lite/images/}.

The latest Raspbian lite bundle can be downloaded from the Raspbian
official webpage~\cite{raspbian}. At the time of this writing, the latest
available bundle was \texttt{2016-05-27-raspbian-jessie-lite.zip}.
To ensure that the content of the bundle does not change, this procedure
is based on that particular version of Raspbian lite which we have
made available at~\cite{tunescode_webpage}. All other files used in this paper
are available there unless specified otherwise. To get started, the testbed
administrator must open a Linux shell (terminal) in and declare the environment
variables shown in the command block below. We show the procedure by
performing the role of the testbed administrator. To copy the commands from
this procedure to the testbed administrator terminal,
we strongly recommend to read this document with Adobe Acrobat Reader in Linux.
Otherwise, copied characters might not render properly.

@@

In this code block, the \texttt{URL} and \texttt{IMAGE} variables specify where
the Linux bundle is located and \texttt{WORKDIR} specifies a working
directory where the Raspbian bundle will be downloaded and customized.
Notice that even though we use the \$ and \# signs in the shell, in general
these signs will be particular to the testbed administrator \ac{OS} shell.
Next, we create the working directory and go to it with the \texttt{cd}
command. To download the image, we utilize the \texttt{wget} command before
unpacking the \texttt{zip} file as follows:

@@

After Raspbian lite has been unpacked, there should be an \texttt{.img}
file in the \texttt{WORKDIR} directory. Here, \texttt{fdisk} can be used to
display the content of the image. To do this and obtain administrative
information, we parse the arguments \texttt{-u sectors} to display the
sizes in sectors and \texttt{-l} to display the partitions within the
image. The \texttt{fdisk} command should output to the terminal something
similar to:

@@

\subsection{Block Devices Mounting}
For browsing and altering the files in the image, we mount the block
device partitions into a particular path of our \texttt{WORKDIR} in order
to customize them. We proceed first to mount the block device partition that
contains the root filesystem and later the boot partition. This is done by
creating an empty directory that it is used as a mountpoint. We name it
\texttt{root} and create it in the working directory before mounting the
root filesystem onto the mountpoint. Thus, we mount the root filesystem
as follows:

@@

The root filesystem mounted in \texttt{ROOTDIR}, has already a boot
directory that can be used as the mount point for the boot partition
in the block device \texttt{loop0p1}. This is convenient because
the final edited partition from \texttt{ROOTDIR/boot} will be mounted on
this same directory when a \ac{Raspi} starts up with a memory card
containing the raw final image. Hence, to mount boot partition we do:
%It is therefore the natural place to mount it for later purposes.

@@

In this way, it is now possible to change all files within the Raspbian
image as desired by editing the files in \texttt{ROOTDIR}. We take
advantage of this to edit configuration files, append new files and even
update and install packages.

@@

\subsubsection{Image Default Configuration Scripts Download}
\label{sec:configuration_files_download}
In our case, we have our default configuration scripts stored in a
\texttt{rasp\_config.zip} file located in the same \ac{URL} where the image
was retrieved from, i.e. the one in the environment variable \texttt{URL}.
We first manually download this compressed file with \texttt{wget} and
extract it locally into our Raspbian lite image. These commands and the
output of the last one are shown as follows:

@@

The unzippped files are one configuration file and three configuration
scripts put in the newly created \texttt{\${ROOTDIR}/home/pi/rasp\_config/}
folder in the image. In what follows, we describe which features do we
require all the \ac{Raspi}s to have and how are they achieved with
these configuration scripts.

@@

Prior to this stage, the \ac{MAC} address of a network card can be found
using the command \texttt{ifconfig} or \texttt{ip addr} on a given
\ac{Raspi}. We store the \ac{MAC} addresses and hostnames of the
\ac{Raspi}s in the configuration file
\texttt{\${ROOTDIR}/home/pi/rasp\_config/nodes.csv}. A sample of our file
is shown as follows:

@@

The testbed administrator has to insert the \ac{MAC} addresses and hostnames
of his/her \ac{Raspi}s obtained previously in the format shown in the
configuration file. In the file format, on each line there is a \ac{MAC}
address and the corresponding hostname for the given \ac{Raspi}. For
simplicity in the example above, we just include three \ac{Raspi}s.
This list file will be employed by the
\texttt{\${ROOTDIR}/home/pi/rasp\_config/set\_hostname}
\ac{Bash} script to assign the hostname of each \ac{Raspi}.
The script content is the following:

>>>>

To get started, we first need to install an \ac{OS} that works properly
on all the \ac{Raspi} models. We will download and setup the image in
the testbed administrator \ac{PC} using a Debian-based distribution. We
use the popular Debian-based Raspbian Linux~\cite{raspbian} given that is
the recommended and default \ac{OS} for the \ac{Raspi}. Raspbian is made
available in two bundles: Raspbian and Raspbian Lite. The difference
between the two, is that Raspbian contains a pre-installed desktop environment
for user interaction and Raspbian Lite by default only permits to interact
through a command shell. Given that the \ac{Raspi}s in our testbed are not
connected to monitors, we decide to work with Raspbian Lite. Still, if later
desired, a desktop environment can be installed using the package manager
as it will be shown in the configuration section.

% The overall procedure to customize the official Raspbian Lite image
% are:
% \begin{enumerate}
%     \item Download Raspbian Lite
%     \item Alter Raspbian Lite. e.g. browse, modify, add, and delete files
%         in the official Raspbian Lite image
%     \item Change root, i.e. change root filesystem into the Raspbian Lite
%         image to update and install software packages
%     \item Write image to memory cards
% \end{enumerate}

%\subsection{Download Raspbian Lite}
%To download the latest Raspbian Lite, we go to the url
%\url{http://downloads.raspberrypi.org/raspbian_Lite/images/}.

The latest Raspbian Lite bundle can be downloaded from the Raspbian
official webpage~\cite{raspbian}. At the time of this writing, the latest
available bundle was \texttt{2016-05-27-raspbian-jessie-lite.zip}.
To ensure that the content of the bundle does not change, this procedure
is based on that particular version of Raspbian Lite which we have
made available at~\cite{tunescode_webpage} acting as our \ac{HTTP} server.
All other files used in this paper are available there unless specified
otherwise. To get started, the testbed administrator must open a Linux
shell (terminal) in its \ac{PC} and declare the environment variables shown
in the command block below. We show the whole procedure by performing the role
of the testbed administrator. To copy the commands from this procedure to the
testbed administrator terminal, we recommend to read this document
with Adobe Acrobat Reader in Linux. Otherwise, copied characters might not
render properly when pasted.

@@

In this code block, the \texttt{\$\{URL\}} and \texttt{\$\{IMAGE\}}
variables specify where the Linux bundle is located and
\texttt{\$\{WORKDIR\}} specifies a working directory where the Raspbian
Lite bundle will be downloaded and customized. Notice that even though we
use the \$ and \# signs in the shell, in general these signs will be
particular to the testbed administrator \ac{OS} shell. Next, we create
the working directory and go to it with the \texttt{cd} command. To
download the image, we utilize the \texttt{wget} command before unpacking
the \texttt{zip} file as follows:

@@

After Raspbian Lite has been unpacked, there should be an \texttt{.img}
file in the \texttt{\$\{WORKDIR\}} directory. Here, \texttt{fdisk} can
be used to display the content of the image. To do this and obtain
administrative information, we parse the arguments \texttt{-u sectors}
to display the sizes in sectors and \texttt{-l} to display the partitions
within the image. The \texttt{fdisk} command should output to the terminal
something similar to:

@@

\subsection{Block Devices Mounting}
For browsing and altering the files in the image, we mount the block
device partitions into a particular path of our \texttt{\$\{WORKDIR\}} in order
to customize them. We proceed first to mount the block device partition that
contains the root filesystem and later the boot partition. This is done by
creating an empty directory that it is used as a mountpoint. We name it
\texttt{root} and create it in the working directory before mounting the
root filesystem onto the mountpoint. Thus, we mount the root filesystem
as follows:

@@

The root filesystem mounted in \texttt{\$\{ROOTDIR\}}, has already a boot
directory that can be used as the mount point for the boot partition
in the block device \texttt{/dev/loop0p1}. This is convenient because
the final edited partition from \texttt{\$\{ROOTDIR\}/boot} will be mounted on
this same directory when a \ac{Raspi} starts up with a memory card
containing the raw final image. Hence, to mount boot partition we do:
%It is therefore the natural place to mount it for later purposes.

@@

In this way, it is now possible to change all files within the Raspbian
image as desired by editing the files in \texttt{\$\{ROOTDIR\}}. We take
advantage of this to edit configuration files, append new files and even
update and install packages.

@@

\subsubsection{Image Default Configuration Scripts Download}
\label{sec:configuration_files_download}
In our case, we have our default configuration scripts stored in a
\texttt{rasp\_config.zip} file located in the same \ac{URL} of the
\ac{HTTP} server where the image was retrieved from, i.e. the one in
the environment variable \texttt{\$\{URL\}}. We first manually download
this compressed file with \texttt{wget} and extract it locally into our
Raspbian Lite image. These commands and the output of the last one are
shown as follows:

@@

The unzippped files are one configuration file and three configuration
scripts put in the newly created \texttt{\$\{ROOTDIR\}/home/pi/rasp\_config/}
folder in the image. In what follows, we describe which features do we
require all the \ac{Raspi}s to have and how are they achieved with
these configuration scripts.

@@

Prior to this stage, the \ac{MAC} address of a network card can be found
using the command \texttt{ifconfig} or \texttt{ip addr} on a given
\ac{Raspi}. We store the \ac{MAC} addresses and hostnames of the
\ac{Raspi}s in the configuration file
\texttt{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}. A sample of our file
is shown as follows:

@@

The testbed administrator has to insert the \ac{MAC} addresses and hostnames
of his / her \ac{Raspi}s obtained previously in the format shown in the
configuration file. In the file format, on each line there is a \ac{MAC}
address and the corresponding hostname for the given \ac{Raspi}. For
simplicity in the example above, we just include three \ac{Raspi}s.
This list file will be employed by the
\texttt{\$\{ROOTDIR\}/home/pi/rasp\_config/set\_hostname}
\ac{Bash} script to assign the hostname of each \ac{Raspi}.
The script content is the following:

### 2016-07-18 introduction.tex
Final submission manuscript

The upcoming 5G technology is targeting the controlling and steering of the Internet of Things (IoT) in real-time on global scale. This will break new ground for new markets such as driverless cars, manufacturing, humanoid robots, smart grids, and many more. With 5G the number of wireless devices will increase by 5$\times$ up to 50 billion devices \cite{cisco2011forecast}. It is generally accepted, that those devices will not be connected in the same cellular manner as the current 10 billion mobile devices are connect due to the expected energy consumption. Therefore the communication architecture in future communication scenario will be dominated by mesh technologies. Mesh technology has been known for sensor and ad hoc networks or mobile cloud scenarios, but the technical requirements on 5G mesh-based communication systems are dramatically increasing. Future mesh networks need to support high data rate, low latency, security, network availability and heterogeneous devices. In state of the art systems those requirements are traded-off with each other, but in the 5G context we cannot do this anymore.

Introduced by Ahlswede et al. \cite{ahlswede2000network}, network coding constitutes a paradigm shift in the way how researchers and industry understand and operate networks, by changing the role of intermediate relays in the process of transmission of information. Relays are no longer limited to storing and forwarding data, but also take part in the coding process, through a process called recoding, where the relay generates new linear combinations of incoming coded packets without previously decoding the data. Network coding allows the increase of throughput, reliability, security and delay performance of the networks. In our previous works, we have shown that \ac{RLNC} \cite{koetter2003algebraic,ho2006random} is able to satisfy the aforementioned technical requirements. We have actually shown how to increase the throughput \cite{pahlevani2013playncool}, reduce the delay \cite{szabo2015towards}, or support heterogeneity for coding enabled communication nodes \cite{lucani2014fulcrum}.

@@

The emergence of powerful and inexpensive single-board computers running full versions of a standard \ac{OS} from many possible opens new possibilities in this area. This not only allows to seamlessly develop implementations that are compatible with higher end devices and allows the reuse stable software from widely-used \ac{OS}. Given this set of specific needs, in this work we present the design, key step-by-step instructions and mechanisms to setup, configurate and maintain of an inexpensive testbed using Raspberry Pi devices for networking (wireless or wired) and storage applications including \ac{RLNC} functionalities into the testbed through Kodo. Nevertheless, the architecture itself is not bounded to the networking area and can be used for other applications that require replicable results with Raspberry Pi. Our work is organized as follows. Section~\ref{sec:testbed} introduces the testbed system, its setup, files configuration and connectivity. Section~\ref{sec:cross_compilation} elaborates on the compilation of the Kodo library for the \ac{Raspi}. Final conclusions and future work are reviewed in Section~\ref{sec:conclusions}.

%The emergence of powerful and inexpensive single-board computers
%running full versions of a standard \ac{OS} from many possible in the last
%years opens exciting possibilities for researchers. This not only
%allows to seamlessly develop implementations that are compatible with
%higher end devices and allows the reuse stable software from the Open
%Source community. It also enables the deployment and testing of large
%scale (hundreds of devices or more) distributed systems at a fraction
%of the cost necessary in previous years. This paper focuses on
%providing the critical steps and mechanisms followed to deploy, setup,
%and maintain a large-scale testbed based on Raspberry Pi devices of
%different models. 
%
%% INTRODUCTION TO NETWORK CODING
%Introduced by Ahlswede et al.~\cite{ahlswede2000network}, network
%coding constitutes a paradigm shift in the way how researchers and
%industry understand and operate networks, by changing the role of
%intermediate relays in the process of transmission of information.
%Relays are no longer limited to storing and forwarding data, but also
%take part in the coding process, through a process called recoding,
%where the relay generates new linear combinations of incoming coded
%packets without previously decoding the data. Network coding allows
%the increase of throughput, reliability, security and delay
%performance of the networks.

%Compared to other broadly used coding schemes, e.g., Gallager's LDPC
%codes~\cite{gallager1962low} or Reed-Solomon
%codes~\cite{reed1960polynomial}, network coding is a technology that
%has been implemented in real systems since the early years of its
%conception. In 2006, Katabi et al.~\cite{katabi2006practical} published
%the results of the performance of their protocol COPE when implemented
%in a real wireless system, which relied on minimalistic coding to
%provide interesting gains. In 2008, Pedersen et
%al.~\cite{pedersen2008implementation} used commercially available
%Symbian OS enabled mobile phones to implement network coding in a
%\ac{D2D} cooperation-based application. In 2011, researchers at Aalborg
%University developed Kodo~\cite{kodo2011pedersen}, a C++11 network
%coding library with open source code for researchers intended to make
%network coding implementations easily available for both, the research
%community and commercial entities.
%Another case is CATWOMAN~\cite{hundeboll2012catwoman}, a protocol
%implemented on top of the BATMAN protocol~\cite{johnson2008simple} for
%IEEE 802.11 multi-hop meshed networks using some of the intuition from
%COPE, but deploying it in real systems. It is currently available as
%open source code in the Linux kernel. Many other implementations have
%been tested on real world systems~
%\cite{pahlevani2013playncool,katti2008xors,krigslund2013core,paramanathan2013leanandmean}.
%
%Despite this successful deployment in real systems, many of these
%protocols and contributions have been implemented in separate testbeds
%and the experiences are hard to reproduce.  Given the need for
%reproducibility of past and future results among researchers and the
%need of low-cost and easy-to-deploy testbeds, this paper contribution
%is provide detailed instructions on how to set up a low-cost,
%heterogeneous and scalable testbed consisting of Raspberry Pis
%(Raspis), and by providing detailed measurements of goodput and energy
%consumption of such testbed when performing network coding operations
%with different codecs (such as full \ac{RLNC} \cite{ho2006random},
%multicore-enabled \ac{RLNC}, sparse \ac{RLNC} and tunable sparse \ac{RLNC}).
%The heterogeneity of the testbed comes from the fact that old, single core,
%Raspberry Pi 1 model B units are integrated seamlessly in the described
%testbed with newer, multicore versions, like the Raspberry Pi 2 model B.
%Our testbed allows us to obtain a comprehensive set of measurements
%to characterize the goodput and energetic footprint of these devices
%showing that processing speeds of 800 Mbps and energy per bit values
%of 1 nJ or even one order of magnitude less are possible.
%
%
%%The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications should be cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the main conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.

>>>>

The upcoming 5G technology is targeting the controlling and steering of the Internet of Things (IoT) in real-time on global scale. This will break new ground for new markets such as driverless cars, manufacturing, humanoid robots, smart grids, and many more. With 5G, the number of wireless devices will increase by 5$\times$ up to 50 billion devices \cite{cisco2011forecast}. It is generally accepted, that those devices will not be connected in the same cellular manner as the current 10 billion mobile devices are connect due to the expected energy consumption. Therefore the communication architecture in future communication scenario will be dominated by mesh technologies. Mesh technology has been known for sensor and ad hoc networks or mobile cloud scenarios, but the technical requirements on 5G mesh-based communication systems are dramatically increasing. Future mesh networks need to support high data rate, low latency, security, network availability and heterogeneous devices. In state of the art systems those requirements are traded-off with each other, but in the 5G context we cannot do this anymore.

Introduced by Ahlswede et al. \cite{ahlswede2000network}, network coding constitutes a paradigm shift in the way how researchers and industry understand and operate networks, by changing the role of intermediate relays in the process of transmission of information. Relays are no longer limited to storing and forwarding data, but also take part in the coding process, through a process called recoding, where the relay generates new linear combinations of incoming coded packets without previously decoding the data. Network coding allows the increase of throughput, reliability, security and delay performance of the networks. In our previous works, we have shown that \ac{RLNC} \cite{koetter2003algebraic,ho2006random} is able to satisfy the aforementioned technical requirements. We have actually shown how to increase the throughput \cite{pahlevani2013playncool}, reduce the delay \cite{szabo2015towards} or support heterogeneity for coding enabled communication nodes \cite{lucani2014fulcrum}.

@@

The emergence of powerful and inexpensive single-board computers running full versions of a standard \ac{OS} from many possible opens new possibilities in this area. This not only allows to seamlessly develop implementations that are compatible with higher end devices and allows the reuse stable software from widely-used \ac{OS}. Given this set of specific needs, in this work we present the design, key step-by-step instructions and mechanisms to setup, configurate and maintain of an inexpensive testbed using Raspberry Pi devices for networking (wireless or wired) and storage applications including \ac{RLNC} functionalities into the testbed through Kodo. Nevertheless, the architecture itself is not bounded to the networking area and can be used for other applications that require replicable results with Raspberry Pi. Our work for the testbed procedure is organized as follows: Section~\ref{sec:overview} introduces the testbed system. In Section~\ref{sec:image_setup}, we provide details about the testbed setup, scripts, configuration files and connectivity. In Section~\ref{sec:overlay_fs}, we elaborate on the need and setup for an overlay filesystem for our testbed in order to have both persistent and non-persistent data on it as an optional step. Section~\ref{sec:tools} describes a set of automation and monitoring tools that can be included in the testbed t simplify the execution of routinary and repetitive tasks. Section~\ref{sec:cross_compilation} elaborates on the compilation of the Kodo library for the \ac{Raspi}. Conclusions and future work are reviewed in Section~\ref{sec:conclusions}. Finally, a set of alternative commands procedures, in case the ones presented in this work might not be executed, are discussed in the Appendix.

%%The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications should be cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the main conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.

### 2016-07-18 overlay_fs.tex
Final submission manuscript

If both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simpy return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian Lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot and will not be modifiable. On the other hand,
to enable the possibility of non-persistent changes, we overlay the upper
filesystem that is mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp}
as read / writable on top of the lower root filesystem. Thus, reading a file
may return a file from the lower filesytem, but if it is stored, it will
be saved in the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be retrieved.

@@

% Get files
\begin{lstlisting}[]
(chroot) # OVERLAYROOTDIR="/tmp/overlayroot"
(chroot) # git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

Although these commands might output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz}, otherwise use \texttt{init-v7.gz} as follows
\todot{What happens if I have both a Raspi 1, 2 or 3 in my testbed? Do I have
to have different images after this point?}:

@@

\subsection{Persistent Image Directories}
\label{sec:persistent_directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are not removed upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent
using the upper filesystem. We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

Again, if the \texttt{----recursive} option is not available, the
procedure in Section~\ref{sec:umount} of the Appendix could be followed.
If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards, although a more
complex setup with a \ac{HTTP} server is also presented in
Section~\ref{sec:http} of the Appendix.

\subsection{Writing Customized Image to Memory Cards}
For a basic system setup, the final step is to write the customized image
to all the memory cards before they can be used in the \ac{Raspi}s. For
our current considered system, we do this manually for each card.
The testbed administrator needs to insert each memory card in its \ac{PC}
and follow the procedure in this section. A given card will be available as
\texttt{/dev/mmcblkX} for memory card readers or as \texttt{/dev/sdbX}
USB-through readers, where \texttt{X} a natural number.

It is \textit{very important} to write to the correct device as
everything will be overwritten. To avoid removing information from the
wrong device, a testbed administrator can insert and eject a memory
card a couple of times and using the \texttt{sudo lsblk} and \texttt{df -h}
commands to deduce the correct device. For our case, the device was
\texttt{/dev/sdb} in an Ubuntu distribution and
\texttt{/dev/mmcblk0} in an Archlinux distribuion. Once identified, to write
the image to a memory card, the following command is used:

@@

The previous \texttt{dd} and \texttt{sync} commands for copying the image to
the SD memory card and flushing the remainder in memory to the filesystem
will take around 15 minutes to be finished. This is an average on a two-core
1.5 GHz laptop. However, other times might occur depending on the total image
size, the writing block size and the core capabilities of the testbed
administrator \ac{PC}. After this is made, it is only required
to eject the memory card and now it can be started in a \ac{Raspi}.

>>>>

If both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simpy return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian Lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot but the local data in these directories will be
erased after a hypothetic reboot. On the other hand, to enable the
possibility of persistent changes, we overlay the upper filesystem that is
mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp} as read / writable
on top of the lower root filesystem. Thus, reading a file may return a
file from the lower filesytem, but if it is stored, it will be saved in
the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be retrieved, but the ones in the lower
filesystem that are not part of the original image will be removed.

@@

% Get files
\begin{lstlisting}[]
(chroot) # OVERLAYROOTDIR="/tmp/overlayroot"
(chroot) # git clone https://github.com/chesty/overlayroot.git ${OVERLAYROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

Although these commands might output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz}, otherwise use \texttt{init-v7.gz} as follows:

@@

\subsection{Persistent and Non-Persistent Image Directories}
\label{sec:non_persistent_directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are removed or not upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent
using the upper filesystem. We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

Again, if the \texttt{----recursive} option is not available, the
procedure in Section~\ref{sec:umount} of the Appendix could be followed.
If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards.

\subsection{Writing Customized Image to SD Memory Cards}
For a basic system setup, the final step is to write the customized image
to all the memory cards before they can be used in the \ac{Raspi}s. For
our current considered system, we do this manually for each card.
The testbed administrator needs to insert each memory card in its \ac{PC}
and follow the procedure in this section. A given card will be available as
\texttt{/dev/mmcblkX} for memory card readers or as \texttt{/dev/sdbX}
USB-through readers, where \texttt{X} a natural number.

It is \textit{very important} to write to the correct device as
everything will be overwritten. To avoid removing information from the
wrong device, a testbed administrator can insert and eject a memory
card a couple of times and using the \texttt{sudo lsblk} or \texttt{df -h}
commands to deduce the correct device. For our case, the device was
\texttt{/dev/sdb} in an Ubuntu distribution and
\texttt{/dev/mmcblk0} in an Archlinux distribution. Once identified, to write
the image to a memory card, the following command is used:

@@

The previous \texttt{dd} and \texttt{sync} commands for copying the image to
the SD memory card and flushing the remainder in memory to the filesystem
will take around 15 minutes to be finished. This is an average on a two-core
1.5 GHz laptop. However, other times might occur depending on the total image
size, the writing block size and the core capabilities of the testbed
administrator \ac{PC}. After this is made, it is only required
to eject the memory card and now plug it in a \ac{Raspi} so it can boot up.
